---
title: "Seminar 2: Introduction to Data I- Solution"
subtitle: "LSE ME314: Introduction to Data Science and Machine Learning"
date-modified: "14 July 2026"
toc: true
format: html
execute:
  echo: true
  eval: true
---

## Plan for Today:
We will learn how to process and manipulate data in R using `tidyverse` and `dplyr`. We will also start learning data visualization using ggplot. 

----

### Recap from Last Seminar:

In your Positron terminal, write the commands below to check that you have installed everything necessary. If any of those commands return an error, go back to yesterday's seminar code and follow the setup instructions.

```{bash, eval=FALSE}

# If you're on macOS or Linux, run:
which R
which git
R --version
quarto check


# If you're on Windows (PowerShell), run:
where R
where git
R --version
quarto check

```


### Recap from last Seminar: Data Types in R

#### 1. Exercise

Create a data frame with three R vectors. In a single R code chunk, write code that:

- Creates a numeric vector called numeric_var containing the values 1,2,3.

- Creates a character vector called character_var containing the strings "data", "code", and "programming".

- Creates a factor called factor_vector from the character vector c("data", "code", "programming").

- Combines the three vectors into a data frame called df using the code: `data.frame(var_name1, var_name2, var_name3) and print the data frame

```{r}
# your code goes here
numeric_var <- c(1, 2, 3)
character_var <- c("data", "code", "programming")
factor_var <- as.factor(character_var)

df <- data.frame(numeric_var, character_var, factor_var)
print(df)
```


## Part I: Processing and Manipulating Data in R

In R, the go-to package for data wrangling and manipulation is `tidyverse` and `dplyr`. Dplyr works with a pipe operator `%>%`. It pipes an object as the first argument into a subsequent function. This is an alternative notation to writing the object as the first input in the function directly.


We will be working with two main packages for this seminar: `dplyr` and `tidyr`. Both come as part of the `tidyverse`, so they are already loaded.

#### Dplyr

Dplyr is the tidyverse package for data manipulation. It offers a collection of functions, so-called verbs, that cover the most common data manipulation tasks for tabular data.

Functions:

- `filter()` and `mutate()`, which creates new variables 

- `group_by()` and `summarise()`, which aggregate values by groups


#### Tidyr

Tidyr is the tidyverse package for reshaping tabular data, such as transforming datasets between wide and long format.

Functions:

- `pivot_wider()`: transforms data from long to wide format

- `pivot_longer()`: transforms data from wide to long format


#### 2. Exercise

Set up your R workspace by installing and loading packages. We will be using the 'tidyverse' package in this session. Import the dataset in the data folder 'ip_and_unemployment.csv' in R using tidyverse.

Install the tidyverse package. You will be asked to select a CRAN mirror which should pop up as a separate window or below in your R terminal. Select the one in UK:London.

```{r, eval=FALSE}
install.packages("tidyverse")
```


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
```


Import the `ip_and_unemployment.csv` dataset using R code, which you find in the data folder of the seminar2 folder using `read.csv()` and call it ip_and_unemployment. Then print out the first rows of the dataset using `head()`

```{r}
ip_and_unemployment <- read.csv("data/ip_and_unemployment.csv")
head(ip_and_unemployment)
```


#### 3. Exercise

Now, let us start working with the dataset to be able to answer the following questions.

What are the highest unemployment rates for France and Spain during the time of the sample?  What are the lowest values for industrial production for the two countries? Make sure to delete NA values in only the time series of interest.

As you could see above when displaying the dataset, it is not in a useful format for us to work with. Therefore, we will change it by following the steps below:

##### Step 1. Pivot the dataset

Start by pivoting the dataset from long to wide format using `pivot_wider()`. You want to turn the `series` column into two separate `unemployment` and `ip` columns and call the new dataset `ipu_clean`.

```{r}
ipu_clean <- ip_and_unemployment %>%
  pivot_wider(
    id_cols = c("country", "date"),
    names_from = "series",
    values_from = "value"
  ) # long to wide
```


##### Step 2. Compute the extrema

For both countries, France and Spain, write separate code and find the maximum unemployment rate and the minimum industrial-production value in the sample using the `filter()` function to filter for the correct country and then filter again to get the highest unemployment rate and the lowest industrial production rate using `max()` and `min()`.

Be sure to drop NA values only when computing each max or min (e.g. via `na.rm = TRUE`), so that other countries or other dates aren’t affected.

Remember that `na.rm = TRUE` applies only to the max()/min() call, so you don’t drop entire rows with missing other variables.

```{r}
# France
ipu_clean %>%
  filter(country == "france") %>%
  filter(
    unemployment == max(unemployment, na.rm = TRUE) |
      ip == min(ip, na.rm = TRUE)
  )
```


```{r}
# Spain
ipu_clean %>%
  filter(country == "spain") %>%
  filter(
    unemployment == max(unemployment, na.rm = TRUE) |
      ip == min(ip, na.rm = TRUE)
  )
```


#### 4. Exercise

Add three new columns to the dataframe using the `mutate()` function

  1. the square of the industrial production percentage change,
  2. the natural logarithm of the unemployment rate, and
  3. the interaction (i.e. the product) of industrial production percentage change and unemployment rate.

```{r}
ipu_clean %>%
  mutate(
    ip_sq = ip^2,
    unemployment_ln = log(unemployment),
    ip_unemployment = ip * unemployment
  ) %>%
  head()
```



Next, we are creating our own tibble, the tidyverse version of a dataframe, to work with it in this exercise.

```{r}

CoffeeOrders <- tibble(
  Product = c(
    "coffee",
    "coffee",
    "coffee",
    "coffee",
    "coffee",
    "latte",
    "latte"
  ),
  Version = c(
    "strong",
    "strong",
    "strong",
    "decaf",
    "decaf",
    "strong",
    "decaf"
  ),
  Size = c("small", "large", "large", "small", "small", "large", "small"),
  Quantity = c(4, 2, 2, 3, 3, 4, 5)
)

print(CoffeeOrders)

```


#### 5. Exercise

Your task is to create a summary table showing the total quantity of each coffee type (product and version) ordered in each size. Use `group_by()` and `summarise()` to aggregate the quantities, and then tidyr's `pivot_wider()` function to transform your data from long format (one row per item) to wide format (coffee types as rows, sizes as columns).

__Hints:__

- `group_by()` should group by the columns containing the coffee types and the sizes

- `summarise()` should be an operation that gives you the total quantity

- `names_from=` should be the column containing the sizes

- `values_from=` should be the column containing the summarised quantities

The results should be a table where each row is a coffee type, each column is a size, and each cell shows the total quantity ordered.

```{r}
OrderSummary <- CoffeeOrders %>%
  group_by(Product, Version, Size) %>% # the rows to group by
  summarise(Quantity = sum(Quantity), .groups = "drop") %>% # the aggregation function to apply
  pivot_wider(
    names_from = Size, # the column that provides the new column names
    values_from = Quantity, # the values to fill the table with
    values_fill = 0 # the value to replace missing values with
  )

print(OrderSummary)
```


#### 6. Exercise (optional)

Import the dataset `ip.csv`, which you find in the data folder, and transform it into long format using `pivot_longer()` and call it 'df_long'. Read the file with `read.csv(..., check.names = FALSE)` so that R does not change the date column names (e.g. `01.01.2019`). The values for the last two months are missing for every country except the US, so drop these missing values when pivoting using `values_drop_na = TRUE`.

```{r}
ip_wide <- read.csv("data/ip.csv", check.names = FALSE) # keep the date column names unchanged
df_long <- ip_wide %>%
  pivot_longer(
    -country,
    names_to = "date",
    values_to = "value",
    values_drop_na = TRUE
  ) # drop the missing last two months
print(df_long)
```


Now, we will mutate the date variable from a character variable to a date object using the `lubridate`package.

```{r}
#install.packages("lubridate")
library(lubridate)

df_long <- df_long %>%
  mutate(date = dmy(date)) # parses "01.01.2019" style strings into real Date objects
```


Lastly, let us examine the dimensions and structure of data objects: We will be using the 'df_long' dataset and inspect the dataset.

```{r}
dim(df_long) # gives you the number of rows and the number of columns
glimpse(df_long) # gives you the data types
summary(df_long$value) # descriptive statistics for one column
summary(df_long) # descriptive statistics for all columns in df_long
count(df_long, country) # count the appearance for each category
```

## Part II: Data Visualization

#### What is visualization?

A visualisation is a **mapping** of **data** to **aesthetics** via visual **geometries** to produce meaning. Formally this can be represented as a "**grammar of graphics**" (Wilkinson, 2005).


#### What is ggplot2?

ggplot2 is Hadley Wickham’s R package for producing “elegant graphics for data analysis” which is based on Lee Wilkinson's Grammar of Graphics.

In `{ggplot2}`, three key pieces:

1. **Data**: The information

2. **Mapping**: Correspondence between data and aesthetics

3. **Layers**: Geometry and transformation

Now, use the data for Italy you generated for exercise 6 from `df_long`and plot it using ggplot: We will first generate a simple line plot using geom_line().

```{r}
# Line plot of results
simple_plot <- ggplot(
  data = df_long %>% filter(country == "italy"),
  aes(x = date, y = value)
) +
  geom_line()

simple_plot
```


Let us add another mapping: we will use the `df_long` for all countries and change the color of the line for each country as well as the size of the line using `linewidth`.

```{r}
# now let us add some aesthetics: we will change the color and thickness of the line

simple_plot_colors <- ggplot(
  data = df_long,
  aes(x = date, y = value, color = country, group = country)
) +
  geom_line(linewidth = 1)

simple_plot_colors
```

Now we can add a theme, a title and change the X and Y axis labels:

```{r}
# now we can add a title and change the X and Y axis labels:

simple_plot_colors +
  labs(
    title = "Industrial Production, 2019-2020",
    x = "Date",
    y = "Industrial Production Value",
    color = "Country"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## By the end of this seminar you should be able to…

- Import data into R

- Manipulate data and datasets and transform them from wide into long format or vice versa or filter based on conditions

- Create new variables using mutate()

- Answer specific research questions by combining multiple data manipulation steps and handle missing values appropriately during calculations

- Visualize distributions and simulations using ggplot2 in R